---
title: 'EDS 223: assignment 4'
author: "Rosemary Juarez"
date: "2022-11-16"
output:
    html_document:
      print_df: paged
      toc: yes
      toc_depth: 4
      toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
getwd()


library(sf)
library(terra)
library(here)
library(dplyr)
library(tmap)
```

## Overview
Marine aquaculture has the potential to play an important role in the global food supply as a more sustainable protein option than land-based meat production.[^hall] [Gentry et al.](https://www.nature.com/articles/s41559-017-0257-9) mapped the potential for marine aquaculture globally based on multiple constraints, including ship traffic, dissolved oxygen, bottom depth .[^gentry]

For this assignment, you are tasked with determining which Exclusive Economic Zones (EEZ) on the West Coast of the US are best suited to developing marine aquaculture for several species of oysters.\

Based on previous research, we know that oysters needs the following conditions for optimal growth:\

-   sea surface temperature: 11-30&deg;C\
-   depth: 0-70 meters below sea level

##### Learning objectives:
-   combining vector/raster data\
-   resampling raster data\
-   masking raster data\
-   map algebra\

### Data
#### Sea Surface Temperature
We will use average annual sea surface temperature (SST) from the years 2008 to 2012 to characterize the average sea surface temperature within the region. The data we are working with was originally generated from [NOAA's 5km Daily Global Satellite Sea Surface Temperature Anomaly v3.1](https://coralreefwatch.noaa.gov/product/5km/index_5km_ssta.php).

#### Bathymetry
To characterize the depth of the ocean we will use the [General Bathymetric Chart of the Oceans (GEBCO)](https://www.gebco.net/data_and_products/gridded_bathymetry_data/#area).[^gebco]

#### Exclusive Economic Zones
We will be designating maritime boundaries using Exclusive Economic Zones off of the west coast of US from [Marineregions.org](https://www.marineregions.org/eez.php).

## Assignment

Below is an outline of the steps you should consider taking to achieve the assignment tasks. 

#### Prepare data (5 points)
To start, we need to load all necessary data and make sure it has the coordinate reference system. 

-   load necessary packages and set path\ 
    -   I recommend using the [`here` package](https://here.r-lib.org/)
-   read in the shapefile for the West Coast EEZ (`wc_regions_clean.shp`)\
-   read in SST rasters
    -   `average_annual_sst_2008.tif`\
    -   `average_annual_sst_2009.tif`\
    -   `average_annual_sst_2010.tif`\
    -   `average_annual_sst_2011.tif`\
    -   `average_annual_sst_2012.tif`\
-   combine SST rasters into a raster stack\
-   read in bathymetry raster (`depth.tif`)\
-   check that data are in the same coordinate reference system\
    -   reproject any data not in the same projection\

```{r include=TRUE, warning=FALSE, messages=FALSE}
#read in West coast EEZ shape file w terra 
wc_EEZ_regions <- st_read(here("data",
                               "wc_regions_clean.shp"))


#read in SST rasters
avg_sst_2008 <- rast(here("data",
                          "average_annual_sst_2008.tif"))
avg_sst_2009 <- rast(here("data",
                          "average_annual_sst_2009.tif"))
avg_sst_2010 <- rast(here("data",
                          "average_annual_sst_2010.tif"))
avg_sst_2011 <- rast(here("data",
                          "average_annual_sst_2011.tif"))
avg_sst_2012 <- rast(here("data",
                          "average_annual_sst_2012.tif"))

#combine into raster stack 
avg_sst <- c(avg_sst_2008,
             avg_sst_2009,
             avg_sst_2010,
             avg_sst_2011,
             avg_sst_2012)

#reading inbathymetry raster
depth <- rast(here("data",
                    "depth.tif"))

#reprojection here
#
#checking all of them first

st_crs(wc_EEZ_regions) #EPSG:4326
st_crs(avg_sst)  #EPSG:9122
st_crs(depth)  #EPSG 4326

#just need to project avg_sst
avg_sst <- project(avg_sst, wc_EEZ_regions)


```


#### Process data (10 points)
Next, we need process the SST and depth data so that they can be combined. In this case the SST and depth data have slightly different resolutions, extents, and positions. We don't want to change the underlying depth data, so we will need to resample to match the SST data using the nearest neighbor approach. 

-   find the mean SST from 2008-2012\
-   convert SST data from Kelvin to Celsius\
    -   hint: subtract by 273.15\
-   crop depth raster to match the extent of the SST raster\
-   note: the resolutions of the SST and depth data do not match\
    -   resample the NPP data to match the resolution of the SST data using the nearest neighbor approach\
-   check that the depth and SST match in resolution, extent, and coordinate reference system\
    -   hint: can the rasters be stacked?
    
```{r include=TRUE}
#find the mean SST from 2008-2012
mean_sst <- mean(avg_sst)

#convert sst from K to C
mean_sst_c <- mean_sst - 273.15

#now im going ot crop depth and mean_sst
depth_crop <- crop(depth,mean_sst_c)


#using nearest neighbor approach to resample
depth_resample <- resample(depth_crop,
                mean_sst_c,
                method = "near")

#stacking right now
resolution_test <- c(depth_resample,
                       avg_sst)

#checking
plot(resolution_test)

```

#### Find suitable locations (20)
In order to find suitable locations for marine aquaculture, we'll need to find locations that are suitable in terms of both SST and depth.

-   reclassify SST and depth data into locations that are suitable for osterz
    -   hint: set suitable values to `1` and unsuitable values to `NA`\
-   find locations that satisfy both SST and depth conditions\
    -   hint: create an overlay using the `lapp()` function multiplying cell values\
    
```{r include=TRUE}
# oysters
# sea surface temperature is 11-30C
# depth is 0-70m below sea level


#reclassification matrix for suitable range of sst
rcl_sst_template <- matrix(c(-Inf, 11, NA, 11, 30,
                    1,30, Inf, NA), 
              ncol = 3, byrow = TRUE) 

#reclassification matrix for suitable range of depths

rcl_depth_template <- matrix(c(-Inf, -70, NA,
                      -70, 0, 1,
                      0, Inf, NA),
                    ncol = 3, byrow = TRUE)
```

```{r}

#reclassifying mean_sst_C
sst_suitable <- classify(mean_sst_c,
                     rcl = rcl_sst_template, 
                     include.lowest = TRUE)


#reclassifying depth
depth_suitable <- classify(depth_resample,
                     rcl = rcl_depth_template,
                     include.lowest = TRUE)
```

```{r}
#This function multiplies two layers, and i got this from the R documentation for lap
fun <- function(x, y) {
  return(x*y)}


#find locations that satisfy both conditions
oyster_suitable <- lapp(c(sst_suitable, depth_suitable), fun = fun)

#checking if it worked
plot(oyster_suitable)
```



#### Determine the most suitable EEZ (20 points)
We want to determine the total suitable area within each EEZ in order to rank zones by priority. To do so, we need to find the total area of suitable locations within each EEZ.

-   select suitable cells within West Coast EEZs\
-   find area of grid cells\
-   find the total suitable area within each EEZ\
    -   hint: it might be helpful to rasterize the EEZ data\
-   find the percentage of each zone that is suitable\
    -   hint it might be helpful to join the suitable area by region onto the EEZ vector data\
    
```{r include=TRUE}
# finidng cell size of areas that are suitable for oysters so we can find the size of each cell
grid_cell_size <- cellSize(oyster_suitable,
                           mask = TRUE,
                           unit = "km")
#checking to see if it worked

plot(grid_cell_size)

```
    

```{r}
# now we need to add the eez zones and rasterize it too, with our newfound suitable oyster location
wc_zones_rast <- rasterize(wc_EEZ_regions,
                           oyster_suitable,
                           transform = TRUE,
                           field = "rgn")
#checking if it worked
plot(wc_zones_rast)
```

```{r}
#make a mask of wc raster and suitable locs for oysters
#goal is to have areas of suitable oyster habitat that also fall in w coast EEZ regions
wc_mask <- mask(wc_zones_rast, oyster_suitable)

plot(wc_mask)
```

```{r}
#now we are setting up the zonal areas with our grid cell and mask!
suitable_zones <- zonal(grid_cell_size, wc_mask, na.rm = TRUE, sum)
```

```{r}
#joining data of the eez regions and suitable zones post-masking
suitable_EEZ <- full_join(wc_EEZ_regions, suitable_zones, by = "rgn") %>%  
  mutate(suitable_area = area,
         percentage_suitable = (suitable_area/area_km2 * 100))

plot(suitable_EEZ)
```


#### Visualize results (5 points)
Now that we have results, we need to present them!

Create the following maps:

-   total suitable area by region\
-   percent suitable area by region\

Include:

-   legible legends\
-   updated color aesthetics\
-   basemap\

```{r include=TRUE}
#map for total suitable area for oysters by region
tm_basemap() +
  tm_shape(suitable_EEZ) +
  tm_polygons('area',
              palette = 'YlGn',
              border.col = 'black',
              title = 'Suitable Area for oysters by region') +
  tm_compass(type="4star", position=c("right", "top")) +
  tm_scale_bar() +
  tm_layout(main.title = 'Total Suitable Area for Oysters in the West Coast EEZ')

#map for percent suitable area by region
tm_basemap() +
  tm_shape(suitable_EEZ) +
  tm_polygons('percentage_suitable',
              palette = 'RdBu',
              border.col = 'black',
              title = 'Percentage of suitable area')+
  tm_layout(main.title = 'Percentage of Suitable Area for Oysters in the West Coast EEZ', main.title.size = 10) +
  tm_compass(type="4star", position=c("right", "top")) +
  tm_scale_bar()
```

#### Broaden your workflow! (40 points)
Now that you've worked through the solution for one group of species, let's update your workflow to work for other species. Please create a function that would allow you to reproduce your results for other species. Your function should be able to do the following:\

-   accept temperature and depth ranges and species name as inputs\
-   create maps of total suitable area and percent suitable area per EEZ with the species name in the title\

Run your function for a species of your choice! You can find information on species depth and temperature requirements on [SeaLifeBase](https://www.sealifebase.ca/search.php). Remember, we are thinking about the potential for marine aquaculture, so these species should have some reasonable potential for commercial consumption. 

```{r}
find_suitable_species <- function(SST_low, SST_high, depth_low, depth_high, species_name) {
  #read in West coast EEZ shape file w terra 
wc_EEZ_regions <- st_read(here("data",
                               "wc_regions_clean.shp"))


#read in SST rasters
avg_sst_2008 <- rast(here("data",
                          "average_annual_sst_2008.tif"))
avg_sst_2009 <- rast(here("data",
                          "average_annual_sst_2009.tif"))
avg_sst_2010 <- rast(here("data",
                          "average_annual_sst_2010.tif"))
avg_sst_2011 <- rast(here("data",
                          "average_annual_sst_2011.tif"))
avg_sst_2012 <- rast(here("data",
                          "average_annual_sst_2012.tif"))

#combine into raster stack 
avg_sst <- c(avg_sst_2008,
             avg_sst_2009,
             avg_sst_2010,
             avg_sst_2011,
             avg_sst_2012)

#reading inbathymetry raster
depth <- rast(here("data",
                    "depth.tif"))

#reprojection here
#
#checking all of them first

st_crs(wc_EEZ_regions) #EPSG:4326
st_crs(avg_sst)  #EPSG:9122
st_crs(depth)  #EPSG 4326

#just need to project avg_sst
avg_sst <- project(avg_sst, wc_EEZ_regions)

  
#find the mean SST from 2008-2012
mean_sst <- mean(avg_sst)

#convert sst from K to C
mean_sst_c <- mean_sst - 273.15

#now im going ot crop depth and mean_sst
depth_crop <- crop(depth,mean_sst_c)


#using nearest neighbor approach to resample
depth_resample <- resample(depth_crop,
                mean_sst_c,
                method = "near")

#stacking right now
resolution_test <- c(depth_resample,
                       avg_sst)


#reclassification matrix for suitable range of sst
rcl_sst_template <- matrix(c(-Inf, SST_low, NA, SST_low, SST_high,
                      1, SST_high, Inf, NA),
                      ncol = 3, byrow = TRUE)

#reclassification matrix for suitable range of depths

rcl_depth_template <- matrix(c(-Inf, depth_low, NA,
                        depth_low, depth_high, 1,
                        depth_high, Inf, NA),
                    ncol = 3, byrow = TRUE)

#reclassifying mean_sst_C
sst_suitable <- classify(mean_sst_c,
                     rcl = rcl_sst_template, 
                     include.lowest = TRUE)


#reclassifying depth
depth_suitable <- classify(depth_resample,
                     rcl = rcl_depth_template,
                     include.lowest = TRUE)
  

#This function multiplies two layers, and i got this from the R documentation for lap
fun <- function(x, y) {
  return(x*y)}


#find locations that satisfy both conditions
oyster_suitable <- lapp(c(sst_suitable, depth_suitable), fun = fun)


# finidng cell size of areas that are suitable for oysters so we can find the size of each cell
grid_cell_size <- cellSize(oyster_suitable,
                           mask = TRUE,
                           unit = "km")


# now we need to add the eez zones and rasterize it too, with our newfound suitable oyster location
wc_zones_rast <- rasterize(wc_EEZ_regions,
                           oyster_suitable,
                           transform = TRUE,
                           field = "rgn")


#make a mask of wc raster and suitable locs for oysters
#goal is to have areas of suitable oyster habitat that also fall in w coast EEZ regions
wc_mask <- mask(wc_zones_rast, oyster_suitable)


#now we are setting up the zonal areas with our grid cell and mask!
suitable_zones <- zonal(grid_cell_size, wc_mask, na.rm = TRUE, sum)

#joining data of the eez regions and suitable zones post-masking
suitable_EEZ <- full_join(wc_EEZ_regions, suitable_zones, by = "rgn") %>%  
  mutate(suitable_area = area,
         percentage_suitable = (suitable_area/area_km2 * 100))

area_species <- tm_basemap() +
  tm_shape(suitable_EEZ) +
  tm_polygons('area',
              palette = 'YlGn',
              border.col = 'black',
              title = 'Suitable Area for',species_name ,'by region') +
  tm_compass(type="4star", position=c("right", "top")) +
  tm_scale_bar() +
  tm_layout(main.title = 'Total Suitable Area for',species_name,'in the West Coast EEZ')

#map for percent suitable area by region
percent_species <- tm_basemap() +
  tm_shape(suitable_EEZ) +
  tm_polygons('percentage_suitable',
              palette = 'RdBu',
              border.col = 'black',
              title = 'Percentage of suitable area')+
  tm_layout(main.title = 'Percentage of Suitable Area for',species_name, 'in the West Coast EEZ', main.title.size = 10) +
  tm_compass(type="4star", position=c("right", "top")) +
  tm_scale_bar()
}

```


```{r, eval = FALSE}
pacific_habitats <- find_suitable_locs(0, 8, 46, 0, "Pacific littleneck")
```


```{r, eval = FALSE}
pacific_habitats
```


[^hall]:Hall, S. J., Delaporte, A., Phillips, M. J., Beveridge, M. & O’Keefe, M. Blue Frontiers: Managing the Environmental Costs of Aquaculture (The WorldFish Center, Penang, Malaysia, 2011). 

[^gentry]:Gentry, R. R., Froehlich, H. E., Grimm, D., Kareiva, P., Parke, M., Rust, M., Gaines, S. D., & Halpern, B. S. Mapping the global potential for marine aquaculture. *Nature Ecology & Evolution*, 1, 1317-1324 (2017).

[^gebco]:GEBCO Compilation Group (2022) GEBCO_2022 Grid (doi:10.5285/e0f0bb80-ab44-2739-e053-6c86abc0289c).


